---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

This document describes some basic analysis of "exercise" data (in terms of number of steps taken per 5-minute interval, over a number of different days).

The data were obtained from Roger Peng's [github](https://github.com/rdpeng/RepData_PeerAssessment1) site as the file "activity.zip", which is then decompressed to produce the input file "activity.csv".   The data can also be obtained directly from [here]("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip").

## Loading and preprocessing the data
Use the `read.csv()` method from R to straightforwardly read the data into a `data.frame` called `activity`:

```{r}
activity <- read.csv("activity.csv")
```

The data contains a number of `NA` fields, however as long as we are aware of this there is no need to pre-process the data any further.

```{r}
sapply(activity, function(x) length(which(is.na(x))))
```

## What is mean total number of steps taken per day?

Once in a `data.frame`, it is straightforward to calculate the number of steps taken per day using `tapply()` with the function `sum`, using the `na.rm = TRUE` argument, and simply taking the mean of the resultant vector

```{r}
stepsByDay <- tapply(activity$steps, activity$date, sum, na.rm = TRUE)
stepsByDay 
                 # The mode, i.e. 'type' of data in R! (not statistical mode)
mode(stepsByDay) # showing that tapply produces a numeric vector in this case
mean(stepsByDay)
median(stepsByDay)
```

The distribution of "steps per day" is best visualized as a histogram

```{r}
hist(stepsByDay, main = "Number of steps per day", xlab = "Steps per day",
     breaks = 30)
```

## What is the average daily activity pattern?
Similarly, using `tapply()`, but this time taking the mean across time intervals (rather than sum across days), we can plot the average number of steps for each time interval throughout the day:

```{r}
stepsByTime <- tapply(activity$steps, activity$interval, mean, na.rm = TRUE)
plot(names(stepsByTime), stepsByTime, type = "l", 
     xlab = "Time Interval (5 min)", ylab = "Number of Steps")
```

Visually inspecting this plot, it appears that the maximum number of steps taken for any 5-minute interval occurs around time interval number 800.

We can determine this exactly:

```{r}
stepsByTime[which(stepsByTime == max(stepsByTime))]
```

which shows that the maximum number of steps (on average) occurs during time interval number 835.

## Imputing missing values
Having calculated the mean number of steps for each time interval throughout the day, it seems reasonable to use this value (for each time interval) to impute missing values.

```{r}
for(day in levels(activity$date)) {
    ind <- which(activity$date == day)
    activity$steps[ind] <-
        ifelse(is.na(activity$steps[ind]), stepsByTime, activity$steps[ind])
}
```

One can check the effect of imputing the missing data by looking at the histogram again:

```{r}
hist(tapply(activity$steps, activity$date, sum), 
     main = "Number of steps per day", xlab = "Steps per day",
     breaks = 30)
```

It is clear that imputing missing values reduces the number of days with "no" steps (number of steps = 0).  Unfortunately this method of imputation also creates a spike at number of steps per day approximately 10,000 where all the days with no data (that were zero steps per day) get the "average" value.

## Are there differences in activity patterns between weekdays and weekends?

Using the `xyplot` function (`lattice` package), with appropriate values for the `panel` and `prepanel` functions:

```{r}
library(lattice)
```

Character vector element of activity for weekend/weekday
```{r}
activity$dayType <- 
    ifelse(weekdays(as.Date(activity$date)) %in% c("Saturday","Sunday"),
           "weekend", "weekday")
```

Convert to factor for use in lattice plots

```{r}
activity$dayType <- as.factor(activity$dayType)
```

Pre-panel function ensures that the y-limits are set correctly
```{r}
myPrepan <- function(x,y) {
    y <- tapply(y, x, mean)
    list(xlim = c(min(x), max(x)),
         ylim = c(min(y),max(y)))
}
```

Panel function does a line plot of the mean (for weekend/weekday)

```{r}
myPanel <- function(x,y) llines(x = x, y = tapply(y, x, mean))

xyplot(steps ~ interval | dayType, data = activity,
       panel = myPanel, prepanel = myPrepan)
```